Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-03-12 numerical-amr.tex
Adding ray tracing method section (a little more than a page).  Fixed a lot of missing references and citations.

Given the grid and it's boundary conditions, the process of solving equations~(\ref{eq:mass_prime})--(\ref{eq:total_energy_prime}) can be detached from the rest of the SAMR technique, so that it is possible to experiment with a number of different methods.  The restrictions are that the method should be flux-conservative, or at least produce an accurate estimate of the flux of conserved quantities at cell interfaces.  This is required in the flux-correction step (see \ref{sec:flux_correction}).  There is also the issue of centering, which refers to the position of physical quantities within a cell (the primary types are cell-centered, face-centered, and node-centered). We will employ both the direct Eulerian version of the piecewise parabolic method \citep{1984JCoPh..54..174C,1995CoPhC..89..149B} which assumes cell-centering for all quantities (actually they are cell-averages).  However, it is also possible to use other schemes, which employ different centering for different variables \citep[e.g.,][]{Stone92a}.   The interpolation step, described in section~\ref{sec:interpolation}, must correctly interpolate with the selected centering.

@@

\item{\em Shear} As described in \citet{KritsukXX}, turbulence calculations can be successfully modeled with AMR if the shear is used as a refinement parameter; in particular, refinement occurs when the finite difference version of the following inequality holds:
\begin{equation}
%\left(\frac{du}{dy}\right)^2 + \left(\frac{du}{dz}\right)^2 + 
%\left(\frac{dv}{dx}\right)^2 + \left(\frac{dv}{dz}\right)^2 + 
%\left(\frac{dw}{dx}\right)^2 + \left(\frac{dw}{dy}\right)^2 
\sum_i \sum_j \left( \frac{\partial v_i}{\partial x_j} \right)^2 -  \sum_i \left( \frac{\partial v_i}{\partial x_i} \right)^2
> \epsilon_s \frac{c_s^2}{\Delta x^2},
\end{equation}
where $\epsilon_s$ is a dimensionless parameter.

@@

>>>>

Given the grid and its boundary conditions, the process of solving
equations~(\ref{eq:mass})--(\ref{eq:total_energy}) can be detached
from the rest of the SAMR technique, so that it is possible to
experiment with a number of different methods.  The restrictions are
that the method should be flux-conservative, or at least produce an
accurate estimate of the flux of conserved quantities at cell
interfaces.  This is required in the flux-correction step (see
\ref{sec:flux_correction}).  There is also the issue of centering,
which refers to the position of physical quantities within a cell (the
primary types are cell-centered, face-centered, and node-centered). We
will employ both the direct Eulerian version of the piecewise
parabolic method \citep{1984JCoPh..54..174C,1995CoPhC..89..149B} which
assumes cell-centering for all quantities (actually they are
cell-averages).  However, it is also possible to use other schemes,
which employ different centering for different variables
\citep[e.g.,][]{Stone92a}.  The interpolation step, described in
section~\ref{sec:interpolation}, must correctly interpolate with the
selected centering.

@@

\item{\em Shear} As described in \citet{Kritsuk06}, turbulence calculations can be successfully modeled with AMR if the shear is used as a refinement parameter; in particular, refinement occurs when the finite difference version of the following inequality holds:
\begin{equation}
%\left(\frac{du}{dy}\right)^2 + \left(\frac{du}{dz}\right)^2 + 
%\left(\frac{dv}{dx}\right)^2 + \left(\frac{dv}{dz}\right)^2 + 
%\left(\frac{dw}{dx}\right)^2 + \left(\frac{dw}{dy}\right)^2 
\sum_i \sum_j \left( \frac{\partial v_i}{\partial x_j} \right)^2 -  \sum_i \left( \frac{\partial v_i}{\partial x_i} \right)^2
> \epsilon_s \frac{c_s^2}{\Delta x^2},
\end{equation}
where $\epsilon_s$ is a dimensionless parameter.

@@

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "ms"
%%% End:

### 2013-03-12 numerical-gravity.tex
Adding ray tracing method section (a little more than a page).  Fixed a lot of missing references and citations.

The dark matter particles are distributed onto the grids using the cloud-in-cell (CIC) interpolation technique to form a spatially discretized density field (analogous to the baryon densities used to calculate the equations of hydrodynamics).  After sampling dark matter density onto the grid and adding baryon density if it exists (to get the total matter density in each cell), the gravitational potential is calculated on the periodic root grid using a fast Fourier transform.  For periodic boundary conditions, we can use either a simple Greens function kernel of $-k^{-2}$, or the finite-difference equivalent \citep{HockneyEastwoord1980}:
\begin{equation}
G(\vec{k}) = - \frac{\Delta x}{2 \left( sin(k_x \Delta x/2)^2 + sin(k_y \Delta y/2)^2 + sin(k_z \Delta z/2)^2 \right) }
\end{equation}
where $k^2 = k_x^2 + k_y^2 + k_z^2$ is the wavenumber in Fourier space and the potential is calculated in k-space as usual with $\tilde{\phi}(k) = G(k) \tilde{\rho}(k)$.

>>>>

The dark matter particles are distributed onto the grids using the cloud-in-cell (CIC) interpolation technique to form a spatially discretized density field (analogous to the baryon densities used to calculate the equations of hydrodynamics).  After sampling dark matter density onto the grid and adding baryon density if it exists (to get the total matter density in each cell), the gravitational potential is calculated on the periodic root grid using a fast Fourier transform.  For periodic boundary conditions, we can use either a simple Greens function kernel of $-k^{-2}$, or the finite-difference equivalent \citep{Hockney88}:
\begin{equation}
G(\vec{k}) = - \frac{\Delta x}{2 \left( sin(k_x \Delta x/2)^2 + sin(k_y \Delta y/2)^2 + sin(k_z \Delta z/2)^2 \right) }
\end{equation}
where $k^2 = k_x^2 + k_y^2 + k_z^2$ is the wavenumber in Fourier space and the potential is calculated in k-space as usual with $\tilde{\phi}(k) = G(k) \tilde{\rho}(k)$.

### 2013-03-12 numerical-overview.tex
Adding ray tracing method section (a little more than a page).  Fixed a lot of missing references and citations.

Instead, \enzo\ uses a particle-mesh N-body method to calculate 
the dynamics of collisionless systems \citep{Hockney88}.  This method 
follows trajectories 
of a representative sample of individual particles and is much more 
efficient than a direct solution of the Boltzmann equation in most 
astrophysical situations. 
The gravitational potential is computed by solving the elliptic 
Poisson's equation (Eqtn.~\ref{enzopoisson}).

@@

We have extended the \enzo\ code to include multiple models for star formation
and feedback.  It is difficult to directly simulate the formation of individual
stars in the context of simulations of galaxy formation and evolution due
to the immense computational cost.  Therefore, 
we adopt parametric methods which attempt
to model star forming regions.  One model is based on Kravtsov's method
 \citep{2003ApJ...590L...1K}, and the other is based on the method of 
\citet{1992ApJ...399L.113C}, which has been modified for use in an AMR code.  The basic
ideas behind the methods are straightforward and observationally motivated.
Similar (though somewhat more advanced) methods have been employed for the 
smoothed particle hydrodynamics
(SPH) method, most recently by Springel \& Hernquist 
\citep{sh03a,sh03b,hs03}. This 
method assumes a
multiphase IGM and has been shown to accurately reproduce the cosmic star
formation rate, and will be implemented into \enzo\ in the near future.

@@

The Cen \& Ostriker method is a heuristic model of star formation
on galactic scales.  This method, first described by
\citet{1992ApJ...399L.113C}, is similar in some ways to the Kravtsov algorithm
but has more complex criteria for determining where stars should
be formed.  In this method, cells that form stars must have a 
baryon overdensity higher than some threshold 
$\rho_b/\bar{\rho}_b \geq \eta$ 
where $\rho_b$ is the baryon density in that cell, 
$\bar{\rho_b}$ is the mean baryon density in the simulation,
and $\eta$ is the user-defined overdensity threshold.
Additionally, the gas in the cells must be contracting, 
cooling rapidly, and gravitationally unstable, e.g.:

>>>>

Instead, \enzo\ uses a particle-mesh N-body method to calculate 
the dynamics of collisionless systems \citep{Hockney88}.  This method 
follows trajectories 
of a representative sample of individual particles and is much more 
efficient than a direct solution of the Boltzmann equation in most 
astrophysical situations. 
The gravitational potential is computed by solving the elliptic 
Poisson's equation (Eq.~\ref{eq:potential}).

@@

We have extended the \enzo\ code to include multiple models for star formation
and feedback.  It is difficult to directly simulate the formation of individual
stars in the context of simulations of galaxy formation and evolution due
to the immense computational cost.  Therefore, 
we adopt parametric methods which attempt
to model star forming regions.  One model is based on Kravtsov's method
 \citep{2003ApJ...590L...1K}, and the other is based on the method of 
\citet{CO1992}, which has been modified for use in an AMR code.  The basic
ideas behind the methods are straightforward and observationally motivated.
Similar (though somewhat more advanced) methods have been employed for the 
smoothed particle hydrodynamics
(SPH) method, most recently by Springel \& Hernquist 
\citep{sh03a,sh03b,hs03}. This 
method assumes a
multiphase IGM and has been shown to accurately reproduce the cosmic star
formation rate, and will be implemented into \enzo\ in the near future.

@@

The Cen \& Ostriker method is a heuristic model of star formation
on galactic scales.  This method, first described by
\citet{CO1992}, is similar in some ways to the Kravtsov algorithm
but has more complex criteria for determining where stars should
be formed.  In this method, cells that form stars must have a 
baryon overdensity higher than some threshold 
$\rho_b/\bar{\rho}_b \geq \eta$ 
where $\rho_b$ is the baryon density in that cell, 
$\bar{\rho_b}$ is the mean baryon density in the simulation,
and $\eta$ is the user-defined overdensity threshold.
Additionally, the gas in the cells must be contracting, 
cooling rapidly, and gravitationally unstable, e.g.:

### 2013-03-12 numerical-radiation-raytracing.tex
Adding ray tracing method section (a little more than a page).  Fixed a lot of missing references and citations.

>>>>

Local radiation from stars and black holes strongly affects their
surroundings.  Radiation transport is a well-studied problem; however,
its treatment in multidimensional calculations is difficult because of
the dependence on seven variables -- three spatial, two angular,
frequency, and time.  The non-local nature of the thermal and
hydrodynamical response to radiation sources further adds to the
difficulty.  The radiative transfer equation in comoving coordinates
\citep{Gnedin97} reads
%
\begin{equation}
  \label{eq:rteqn}
  \frac{1}{c} \; \frac{\partial I_\nu}{\partial t} + 
  \frac{\hat{n} \cdot \nabla I_\nu}{\bar{a}} -
  \frac{H}{c} \; \left( \nu \frac{\partial I_\nu}{\partial \nu} -
  3 I_\nu \right) = -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
Here $I_\nu \equiv I(\nu, \mathbf{x}, \Omega, t)$ is the radiation
specific intensity in units of energy per time $t$ per solid angle per
unit area per frequency $\nu$.  $H = \dot{a}/a$ is the Hubble
constant, where $a$ is the scale factor.  $\bar{a} = a/a_{em}$ is the
ratio of scale factors at the current time and time of emission.  The
second term represents the propagation of radiation, where the factor
$1/\bar{a}$ accounts for cosmic expansion.  The third term describes
both the cosmological redshift and dilution of radiation.  On the
right hand side, the first term considers the absorption coefficient
$\kappa_\nu \equiv \kappa_\nu(\mathbf{x},\nu,t)$, and the second term
$j_\nu \equiv j_\nu(\mathbf{x},\nu,t)$ is the emission coefficient
that includes any point sources of radiation or diffuse radiation.

We can make some appropriate approximations to reduce the complexity
of Equation (\ref{eq:rteqn}) in order to include radiation transport
in numerical calculations.  Typically timesteps in dynamic
calculations are small enough so that $\Delta a/a \ll 1$, therefore
$\bar{a} = 1$ in any given timestep, reducing the second term to
$\hat{n} \partial I_\nu/\partial \mathbf{x}$.  To determine the
importance of the third term, we evaluate the ratio of the third term
to the second term.  This is $HL/c$, where $L$ is the simulation box
length.  If this ratio is $\ll 1$, we can ignore the third term.  For
example at $z=5$, this ratio is 0.1 when $L = c/H(z=5)$ = 53 proper
Mpc.  In large boxes where the light crossing time is comparable to
the Hubble time, then it becomes important to consider cosmological
redshifting and dilution of the radiation.  Thus equation
(\ref{eq:rteqn}) reduces to the non-cosmological form in this local
approximation,
%
\begin{equation}
  \frac{1}{c} \frac{\partial I_\nu}{\partial t} + 
  \hat{n} \frac{\partial I_\nu}{\partial \mathbf{x}} =
  -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
We choose to represent the source term $j_\nu$ as point sources of
radiation (e.g. stars, quasars) that emit radial rays that are
propagated along the direction $\hat{n}$.

Ray tracing is an accurate method to propagate radiation from point
sources on a computational grid, given that there are a sufficient
number of rays passing through each cell.  Along a ray, the radiation
transfer equation reduces to
%
\begin{equation}
\label{eqn:rtray}
\frac{1}{c} \frac{\partial P}{\partial t} + \frac{\partial P}{\partial
  r} = -\kappa P,
\end{equation}
where $P$ is the photon number flux along the ray.  To sample the
radiation field at large radii, ray tracing requires at least $N_{ray}
= 4\pi R^2 / (\Delta x)^2$ rays to sample each cell with one ray,
where $R$ is the radius from the source to the cell and $\Delta x$ is
the cell width.  If one were to trace $N_{ray}$ rays out to $R$, the
cells at a smaller radius $r$ would be sampled by, on average,
$(r/R)^2$ rays, which is computationally wasteful because only a few
rays per cell, as we will show later, provides an accurate calculation
of the radiation field.

We avoid this inefficiency by utilizing adaptive ray tracing
\citep{Abel02_RT}, which progressively splits rays when the sampling
becomes too coarse and is based on Hierarchical Equal Area isoLatitude
Pixelation \citep[HEALPix;][]{HEALPix}.  In this scheme, the rays are
traced along normal directions of the centers of HEALPix pixels, which
evenly divides a sphere into equal areas.  The rays are initialized at
each point source with the photon luminosity (ph s$^{-1}$) equally
spread across $N_{\rm pix} = 12 \times 4^l$ rays, where $l$ is the
initial HEALPix level.  We usually find $l$ = 0 or 1 is sufficient
because these coarse rays will usually be split before traversing
the first cell.

The rays are traced through the grid in a typical fashion
\citep[e.g.][]{Abel99_RT}, in which we calculate the next cell
boundary crossing.  The ray segment length crossing the cell is
%
\begin{equation}
  \label{eqn:trace}
  dr = R_0 - \min_{i=1 \rightarrow 3} \left[(x_{\rm cell,i} - x_{\rm src,i}) /
    \hat{n}_{\rm ray,i} \right],
\end{equation}
%
where $R_0$, $\hat{n}_{\rm ray}$, $x_{\rm cell,i}$, and $x_{\rm
  src,i}$ are the initial distance traveled by the ray, normal
direction of the ray, the next cell boundary crossing in the $i$-th
dimension, and the position of the point source that emitted the ray,
respectively.  However before the ray travels across the cell, we
evaluate the ratio of the face area $A_{\rm cell}$ of the current cell
and the solid angle $\Omega_{\rm ray}$ of the ray,
%
\begin{equation}
  \label{eqn:split}
  \Phi_c = \frac{A_{\rm cell}} {\Omega_{\rm ray}} = 
  \frac{N_{\rm pix} (\Delta x)^2} {4\pi R_0^2}.
\end{equation}
%
If $\Phi_c$ is less than a pre-determined value (usually $>3$), the
ray is split into 4 child rays.  The pixel numbers of the child rays
$p^\prime$ are given by the ``nested'' scheme of HEALPix at the next
level, i.e. $p^\prime = 4 \times p + [0,1,2,3]$, where $p$ is the
original pixel number.  The child rays (1) acquire the new normal
vectors of the pixels, (2) retain the same radius of the parent ray,
and (3) gets a quarter of the photon flux of the parent ray.
Afterward the parent ray is discontinued.

A ray propagates and splits until (1) the photon has traveled $c
\times dt_P$, where $dt_P$ is the radiative transfer timestep; (2) its
photon flux is almost fully absorbed ($>99.9\%$) in a single cell,
which significantly reduces the computational time if the radiation
volume filling fraction is small; (3) the photon leaves the
computational domain with isolated boundary conditions; or (4) the
photon travels $\sqrt{3}$ of the simulation box length with periodic
boundary conditions.

In the first case, the photon is halted at that position and saved,
where it will be considered in the solution of $I_\nu$ at the next
timestep.  In the next timestep, the photon will encounter a different
hydrodynamical and ionization state, hence $\kappa$, in its path.
Furthermore any time variations of the luminosities will be retained
in the radiation field.  This is how this method retains the time
derivative of the radiative transfer equation.  The last restriction
prevents our method from considering sources external to the
computational domain, but a uniform radiation background can be used
in conjunction with ray tracing that adds the local radiation field to
the background intensity.

The radiation field is calculated by integrating equation
(\ref{eqn:rtray}) along each ray, which is done by considering the
discretization of the ray into segments.  In the following
description, we assume the rays are monochromatic for simplicity.  For
convenience, we express the integration in terms of optical depth
$\tau = \int \kappa(r,t) \; dr$, and for a ray segment,
%
\begin{equation}
  \label{eqn:dtau}
  d\tau = \sigma_{\rm abs}(\nu) n_{\rm abs} dr.
\end{equation}
Here $\sigma_{\rm abs}$ and $n_{\rm abs}$ are the cross section and
number density of the absorbing medium, respectively.  In the static
case, equation (\ref{eqn:rtray}) has a simple exponential analytic
solution, and the photon flux of a ray is reduced by
%
\begin{equation}
  \label{eqn:flux}
  dP = P \times (1 - e^{-\tau})
\end{equation}
as it crosses a cell.  We equate the photo-ionization rate to the
absorption rate, resulting in photon conservation \citep{Abel99_RT,
  Mellema06}.  Thus the photo-ionization $k_{\rm ph}$ and
photo-heating $\Gamma_{\rm ph}$ rates associated with a single ray are
%
\begin{equation}
  \label{eqn:kph}
  k_{\rm ph} = \frac{P (1 - e^{-\tau})}{n_{\rm abs} \; V_{\rm cell} \; dt_P},
\end{equation}
\begin{equation}
  \label{eqn:gamma}
  \Gamma_{\rm ph} = k_{\rm ph} \; (E_{\rm ph} - E_i),
\end{equation}
where $V_{\rm cell}$ is the cell volume, $E_{\rm ph}$ is the photon
energy, and $E_i$ is the ionization energy of the absorbing material.
In each cell, the photo-ionization and photo-heating rates from each
ray in the calculation are summed, and after the ray tracing is
complete, these rates are used to update the ionization state and
energy of the cells with the solver described in \S\ref{sec.ov.chem}.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "ms"
%%% End:

### 2013-03-12 overview.tex
Adding ray tracing method section (a little more than a page).  Fixed a lot of missing references and citations.

%\begin{eqnarray} 
%\frac{\partial \rho}{\partial t} 
%          + \vecv \cdot \nabla \rho
%    & = & - \rho \nabla \cdot \vecv,
%        \label{eq:mass} \\
%%
%\frac{\partial \vecv }{ \partial t} 
%          + ( \vecv \cdot \nabla ) \vecv 
%    & = & - \frac{1}{\rho} \nabla p
%          - \nabla \phi,
%        \label{eq:momentum} \\
%%
% \frac{\partial E}{\partial t} + \vecv \cdot \nabla E
%    & = & - \frac{1}{\rho} \nabla \cdot (p \vecv) 
%          - \vecv \cdot \nabla \phi
%%\nonumber          \\  & &     % uncomment for 2-column
%     + \Gamma - \Lambda
%        \label{eq:total_energy}
%\end{eqnarray}
%
Here $E$, $e$, $\rho$, $\vecv$ and $p$ are the comoving total specific fluid energy, specific thermal energy, baryonic density, velocity and pressure, respectively.  
%Here we use primes to indicate variables defined in the absolute frame of reference (to distinguish them from a comoving frame).  
The first equation represents conservation of mass, the second, conservation of momentum, and the third, conservation of total fluid (kinetic plus thermal) energy.  They are respectively, the first, second and third moments of the Boltzmann equation. 
Terms representing radiative cooling ($\Lambda$) and heating ($\Gamma$) enter on the right-hand side of the energy equation (\ref{eq:total_energy_prime}).

@@

\subsubsection{Hydrodynamics: PPM method}

@@

\subsubsection{Hydrodynamics: ZEUS method}

>>>>

%\begin{eqnarray} 
%\frac{\partial \rho}{\partial t} 
%          + \vecv \cdot \nabla \rho
%    & = & - \rho \nabla \cdot \vecv,
%        \label{eq:mass} \\
%%
%\frac{\partial \vecv }{ \partial t} 
%          + ( \vecv \cdot \nabla ) \vecv 
%    & = & - \frac{1}{\rho} \nabla p
%          - \nabla \phi,
%        \label{eq:momentum} \\
%%
% \frac{\partial E}{\partial t} + \vecv \cdot \nabla E
%    & = & - \frac{1}{\rho} \nabla \cdot (p \vecv) 
%          - \vecv \cdot \nabla \phi
%%\nonumber          \\  & &     % uncomment for 2-column
%     + \Gamma - \Lambda
%        \label{eq:total_energy}
%\end{eqnarray}
%
Here $E$, $e$, $\rho$, $\vecv$ and $p$ are the comoving total specific fluid energy, specific thermal energy, baryonic density, velocity and pressure, respectively.  
%Here we use primes to indicate variables defined in the absolute frame of reference (to distinguish them from a comoving frame).  
The first equation represents conservation of mass, the second, conservation of momentum, and the third, conservation of total fluid (kinetic plus thermal) energy.  They are respectively, the first, second and third moments of the Boltzmann equation. 
Terms representing radiative cooling ($\Lambda$) and heating ($\Gamma$) enter on the right-hand side of the energy equation (\ref{eq:total_energy}).

@@

\subsubsection{Godunov PPM method (Hydrodynamics)}

@@

\subsubsection{Godonuv MUSCL with Dednet (HD and MHD)}

\subsubsection{MHDCT}

\subsubsection{Second order finite difference method (HD)}

### 2013-03-12 parallel.tex
Adding ray tracing method section (a little more than a page).  Fixed a lot of missing references and citations.

Because of the wide variety of simulations, methods, and uses of \bb{enzo}, it
is difficult to define exactly which routines are most costly during a given
simulation.  As such, we have designed a lightweight registering system that
has been implemented for the most commonly used routines (such as the
hydrodynamic and gravity solvers) as well as refinement level timers that
measure the (exclusive) time spent on each level.  Beyond this minimal set of
routines, we have designed a simple way for the user to modify the source
by adding TIMER_START(``Your Routine Name'') and TIMER_END(``Your Routine
Name'').  These timers are automatically registered in a
std::map\footnote{http://www.cplusplus.com/reference/map/map/}.  These timers
are created and managed individually on each processor in an asynchronous
fashion.

>>>>

Because of the wide variety of simulations, methods, and uses of Enzo,
it is difficult to define exactly which routines are most costly
during a given simulation.  As such, we have designed a lightweight
registering system that has been implemented for the most commonly
used routines (such as the hydrodynamic and gravity solvers) as well
as refinement level timers that measure the (exclusive) time spent on
each level.  Beyond this minimal set of routines, we have designed a
simple way for the user to modify the source by adding
\texttt{TIMER\_START(``Your Routine Name'')} and
\texttt{TIMER\_END(``Your Routine Name'')}.  These timers are
automatically registered in a
std::map\footnote{http://www.cplusplus.com/reference/map/map/}.  These
timers are created and managed individually on each processor in an
asynchronous fashion.
